---
title: "Data Exploration and Preparation"
author: 
  - name: Florian J. Auer
    url: https://github.com/florian-j-auer
    affiliation: IT-Infrastructure for Translational Medical Research, University of Augsburg
    affiliation_url: https://www.uni-augsburg.de/de/fakultaet/fai/informatik/prof/misit/
date: "03/30/2022"
output: 
  BiocStyle::html_document:
    css: custom.css
    self_contained: no
    toc: yes
    toc_float: yes
  BiocStyle::pdf_document:
    toc: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=55,
                                     args.newline = TRUE,
                                     arrow = TRUE),
                      tidy=TRUE)
```

```{r navigation, eval=HTML, results='asis', echo=FALSE}
res <- knitr::knit_child('navigation/navigation.Rmd', quiet = TRUE)
cat(res, sep = '\n')
```


# ndexr

Installation via Bioconductor:

```{r ndexrBioconductor, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("ndexr")

library(ndexr)
```


# RCX

For installing packages from github the `devtools` package is the most common approach.
However, it requires XML libraries installed on the system which can cause problems while installation due to unmet dependencies.
The `remotes` package covers the functionality to download and install R packages stored in 'GitHub', 'GitLab', 'Bitbucket', 'Bioconductor', or plain 'subversion' or 'git' repositories without depending on XML libraries.
If `devtools` is already installed, of course it can be used, otherwise it is recommended to use the lightweight `remotes` package.


## From Bioconductor:

```{r rcxBioconductor, eval=FALSE}
if(!"BiocManager" %in% installed.packages()){
  install.packages("BiocManager")
}
if(!"RCX" %in% installed.packages()){
  BiocManager::install("RCX")
}
library(RCX)
```

## From github using remotes:

```{r rcxGithub, eval=FALSE}
if(!"remotes" %in% installed.packages()){
  install.packages("remotes")
}
if(!"RCX" %in% installed.packages()){
  remotes::install_github("frankkramer-lab/RCX")
}
library(RCX)
```

## From github using devtools:

```{r rcxDevtools, eval=FALSE}
if(!"devtools" %in% installed.packages()){
  install.packages("devtools")
}
if(!"RCX" %in% installed.packages()){
  devtools::install_github("frankkramer-lab/RCX")
}
library(RCX)
```




```{r navigation2, eval=HTML, results='asis', echo=FALSE}
res <- knitr::knit_child(HTML_file, quiet = TRUE)
cat(res, sep = '\n')
```

# Session info
```{r sessionInfo}
sessionInfo()
```

